---
title: "IPEDS Financial Metrics"
author: "Jason P. Casey"
format:
  revealjs:
    self-contained: true
    footer: "[Institutional Effectiveness and Analytics](https://iea.unl.edu);"
    logo: Nebraska_N_RGB.png
    scrollable: true
    theme: custom.scss
    preview-links: auto
    slide-number: c/t
execute: 
  echo: false
editor: source
---

```{r setup}
#| label: load-packages
#| include: false
library(purrr, warn.conflicts = F)
library(glue, warn.conflicts = F)
library(janitor, warn.conflicts = F)
library(lubridate, warn.conflicts = F)
library(ggbeeswarm, warn.conflicts = F)
library(knitr, warn.conflicts = F)
library(kableExtra, warn.conflicts = F)
library(formattable, warn.conflicts = F)
library(data.table, warn.conflicts = F)
library(tidyverse, warn.conflicts = F)

options(knitr.kable.NA = '')
options(tidyverse.quiet = TRUE)
options(dplyr.summarise.inform = FALSE)

# set random seed
set.seed(1965)
```


```{r}
#| label: local-functions
#| include: false

# create year keys for fetching data
fetch_years <- function(window_years = 10) {
# set constants
  last_year <- ifelse(month(today()) > 9,
                        year(today()) - 1,
                        year(today()) - 2)
  
  # base tibble used for all data pulls
  # because the window includes the final year, it must be offset by one
  years <- data.table(year_key = seq(last_year - window_years + 1, last_year))
  years[, `Fiscal Year` := paste(year_key - 1, year_key, sep = '-')]
  return(years)
}


fetch_gdp <- function(years) {
  gdp <-
    fread('https://fred.stlouisfed.org/graph/fredgraph.csv?id=GDPDEF',
          colClasses = 'character',
          strip.white = T) |>
    clean_names()
  
  gdp[, date := as_date(date)]
  
  gdp[month(date) == 4,
      .(year_key = ifelse(month(date) > 6, year(date) + 1, year(date)),
        gdp = as.double(gdpdef) / 100)][year_key %in% years]
}

table_me <- function(df) {
  df |>
    kbl() |>
    kable_paper(c('condensed', 'hover'), full_width = F, position = 'center')
}

# include user functions
source(file = "user_functions.R")

unnest_dt <- function(dt, col, id = list(year_key)){
  stopifnot(is.data.table(dt))
  
  by <- substitute(id)
  col <- substitute(unlist(col, recursive = F))
  
  dt[, eval(col), by = eval(by)]
}

fetch_hd <- function(year_key) {
  inst <- load_hd(year_key)[iclevel == 1 & obereg < 9 & opeflag == 1 &
                              deggrant == 1 & act == 'A' & control %in% 1:2]
  ic <- load_ic(year_key)[distnced == 2 & ft_ug == 1, .(unitid)]
  setkey(inst, unitid)
  setkey(ic, unitid)
  ic[inst, nomatch = NULL][,
           .(unitid, institution_name = instnm,  state = stabbr, fips,
             longitude = longitud, latitude)]
}

fetch_fall_enrollment <- function(year_key) {
  load_fall_enrollment(year_key)
}

fetch_student_aid <- function(year_key) {
  df <- load_sfa(year_key)[field %in% c('agrnt_t', 'igrnt_t', 'igrnt_p',
                                        'loan_p', 'loan_a', 'pgrnt_p')]
  df <- dcast(df,
              unitid ~ field,
              value.var = 'value')
  
  df[, .(unitid,
         total_grant_aid = as.integer(agrnt_t),
         inst_grant_aid = as.integer(igrnt_t),
         pct_inst_grant_aided = igrnt_p / 100,
         pct_pell_aided = pgrnt_p / 100,
         total_loan_amount = as.integer(loan_a),
         pct_with_loans = loan_p/ 100)]
}

fetch_submissions <- function(year_key) {
  load_submissions(year_key)[
    ,
    .(unitid,
      parent_id = ifelse(prch_f %in% c('1', '2', '3'), idx_f, unitid))]
}

fetch_retention <- function(year_key) {
  df <- load_retention(year_key)
  df[, .(unitid,
         ft_retention_rate = ret_pcf / 100,
         pt_retention_rate = ret_pcp / 100)]
}

fetch_charges <- function(year_key) {
  df <- load_charges(year_key)[field %in% c('tuition2', 'tuition3')]
  df <- dcast(df,
              unitid ~ field,
              value.var = 'value')
  df[,
     .(unitid,
       in_state_tuition = tuition2,
       out_of_state_tuition = tuition3)]
}

fetch_finance <- function(year_key) {
  fasb <- load_fasb(year_key)[
    field %in% c('f2b01', 'f2b02', 'f2d01', 'f2a03a', 'f2a17', 'f2a18', 'f2c05',
                 'f2c06', 'f2c08', 'f2c10',  'f2d01', 'f2e135', 'f2h02'),
    accounting_standard := 'FASB']
  
  fasb[, field := recode(field,
                         'f2b01' = 'Total Revenues',
                         'f2b02' = 'Total Expenses',
                         'f2a03a' = 'Total Debt',
                         'f2a17' = 'Total Plant',
                         'f2a18' = 'Accumulated Depreciation',
                         'f2c05' = 'Institutional Grants (Funded)',
                         'f2c06' = 'Institutional Grants (Unfunded)',
                         'f2c08' = 'Allowances to Tuition and Fees',
                         'f2c10' = 'Total Discounts',
                         'f2d01' = 'Tuition Revenue',
                         'f2e135' = 'Depreciation Expense',
                         'f2h02' = 'Endowment EOY')]
  
  gasb <- load_gasb(year_key)[
    field %in% c('f1d01', 'f1d02', 'f1a07', 'f1a27t4', 'f1a284', 'f1e05',
                 'f1e06', 'f1e08', 'f1e10', 'f1b01', 'f1c19dp', 'f1h02',
                 'f1b09'),
    accounting_standard := 'GASB']
  gasb[, field := recode(field,
                         'f1d01' = 'Total Revenues',
                         'f1d02' = 'Total Expenses',
                         'f1a07' = 'Total Debt',
                         'f1a27t4' = 'Total Plant',
                         'f1a284' = 'Accumulated Depreciation',
                         'f1e05' = 'Institutional Grants (Funded)',
                         'f1e06' = 'Institutional Grants (Unfunded)',
                         'f1e08' = 'Allowances to Tuition and Fees',
                         'f1e10' = 'Total Discounts',
                         'f1b01' = 'Tuition Revenue',
                         'f1c19dp' = 'Depreciation Expense',
                         'f1h02' = 'Endowment EOY')]
  df <- funion(fasb, gasb)

  dcast(df,
        unitid + accounting_standard ~ field,
        value.var = 'amount')
}
```

## Data Sources

The underlying data for this presentation comes from various public sources:

- *GDP*: Gross Domestic Product from the Federal Reserve
- *Fall Enrollment*: IPEDS Fall Enrollment (EF) Survey, Part A
- *Fall Enrollment by State*: IPEDS Fall Enrollment (EF) Survey, Part C
_ *First-year Retention*: IPEDS Fall Enrollment (EF) Survey, Part D

```{r}
#| label: load-general-data
#| include: false

# EF headcount FTE weights
# see 'Calculation of FTE students (using fall student heacounts) in
#    https://surveys.nces.ed.gov/ipeds/VisGlossaryAll.aspx
weights <- data.table(time_status = rep(c('Full-time', 'Part-time'),
                                        times = 1, each = 24),
                      career_level = rep(c('Undergraduate', 'Graduate'),
                                         times = 6, each = 4),
                      Control = rep(c('Public', 'Private', 'For-profit'),
                                    times = 2, each = 8),
                      Level = c(1, 2, 3, -3),
                      weight = c(rep(1, 24),
                                 c(0.403543, 0.335737, 0.335737, 0.382059,
                                   0.361702, 0.335737, 0.335737, 0.335737,
                                   0.392857, 0.335737, 0.335737, 0.335737,
                                   0.382059, 0.335737, 0.335737, 0.335737,
                                   0.392857, 0.335737, 0.335737, 0.335737,
                                   0.382059, 0.335737, 0.335737, 0.335737))
                      )
years <- fetch_years(10)
gdp <- fetch_gdp(years$year_key)
institutions <- copy(years)[, data := map(year_key, fetch_hd)]
submissions <- copy(years)[, data := map(year_key, fetch_submissions)]
# directory <- unnest_dt(directory, data, list(year_key))
fall_enrollment <- copy(years)[, data := map(year_key, fetch_fall_enrollment)]
student_aid <- copy(years)[, data := map(year_key, fetch_student_aid)]
retention <- copy(years)[, data := map(year_key, fetch_retention)]
charges <- copy(years)[, data := map(year_key, fetch_charges)]
finance <- copy(years)[, data := map(year_key, fetch_finance)]
```

## Years Covered

::: {style="font-size: 0.65em"}
For IPEDS Survey components, fiscal years and term data (e.g., Fall Enrollment) are offset by one year in a collection cycle.  For our purposes, this offset is not much of a problem, so there has been no correction (i.e., aligning a fall term with its correct fiscal cycle).
:::

::: {style="font-size: 0.75em"}
```{r}
#| label: show-years
years |>
  mutate(`Term Data` = paste0('Fall ', year_key)) |>
  rename(`Collection Year` = year_key) |>
  table_me()
```
:::

## Test

```{r}
unnest_dt(finance, data)
```

