---
title: "IPEDS Financial Metrics"
subtitle: "NCES Data Files"
author: "Jason P. Casey"
date: 2020-04-03
output: html_notebook
params:
  year:
    value: 2018
---

```{r setup}
# libraries
library(knitr)
library(DBI)
library(odbc)
library(dbplyr)
library(RSQLite)
library(lubridate)
library(broom)
library(tidyverse)

# set default document options
knitr::opts_chunk$set(fig.width = 8,
                      fig.asp = 0.618,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      collapse = TRUE)

# set random seed
set.seed(1965)

# capture year_param for queries
year_param <- params$year

# EF headcount FTE weights
# see 'Calculation of FTE students (using fall student heacounts) in
#    https://surveys.nces.ed.gov/ipeds/VisGlossaryAll.aspx
weights <-
  tribble(~time_status, ~career_level, ~Control, ~Level, ~weight,
          # full-time weights are all 1.0
          'Full-time', 'Undergraduate', 'Public', 1, 1.0,
          'Full-time', 'Undergraduate', 'Public', 2, 1.0,
          'Full-time', 'Undergraduate', 'Public', 3, 1.0,
          'Full-time', 'Undergraduate', 'Public', -3, 1.0,
          'Full-time', 'Graduate', 'Public', 1, 1.0,
          'Full-time', 'Graduate', 'Public', 2, 1.0,
          'Full-time', 'Graduate', 'Public', 3, 1.0,
          'Full-time', 'Graduate', 'Public', -3, 1.0,
          'Full-time', 'Undergraduate', 'Private', 1, 1.0,
          'Full-time', 'Undergraduate', 'Private', 2, 1.0,
          'Full-time', 'Undergraduate', 'Private', 3, 1.0,
          'Full-time', 'Undergraduate', 'Private', -3, 1.0,
          'Full-time', 'Graduate', 'Private', 1, 1.0,
          'Full-time', 'Graduate', 'Private', 2, 1.0,
          'Full-time', 'Graduate', 'Private', 3, 1.0,
          'Full-time', 'Graduate', 'Private', -3, 1.0,
          'Full-time', 'Undergraduate', 'For-profit', 1, 1.0,
          'Full-time', 'Undergraduate', 'For-profit', 2, 1.0,
          'Full-time', 'Undergraduate', 'For-profit', 3, 1.0,
          'Full-time', 'Undergraduate', 'For-profit', -3, 1.0,
          'Full-time', 'Graduate', 'For-profit', 1, 1.0,
          'Full-time', 'Graduate', 'For-profit', 2, 1.0,
          'Full-time', 'Graduate', 'For-profit', 3, 1.0,
          'Full-time', 'Graduate', 'For-profit', -3, 1.0,
          # part-time
          'Part-time', 'Undergraduate', 'Public', 1, 0.403543,
          'Part-time', 'Undergraduate', 'Public', 2, 0.335737,
          'Part-time', 'Undergraduate', 'Public', 3, 0.335737,
          'Part-time', 'Undergraduate', 'Public', -3, 0.382059,
          'Part-time', 'Graduate', 'Public', 1, 0.361702,
          'Part-time', 'Graduate', 'Public', 2, 0.335737,
          'Part-time', 'Graduate', 'Public', 3, 0.335737,
          'Part-time', 'Graduate', 'Public', -3, 0.335737,
          'Part-time', 'Undergraduate', 'Private', 1, 0.392857,
          'Part-time', 'Undergraduate', 'Private', 2, 0.335737,
          'Part-time', 'Undergraduate', 'Private', 3, 0.335737,
          'Part-time', 'Undergraduate', 'Private', -3, 0.335737,
          'Part-time', 'Graduate', 'Private', 1, 0.382059,
          'Part-time', 'Graduate', 'Private', 2, 0.335737,
          'Part-time', 'Graduate', 'Private', 3, 0.335737,
          'Part-time', 'Graduate', 'Private', -3, 0.335737,
          'Part-time', 'Undergraduate', 'For-profit', 1, 0.392857,
          'Part-time', 'Undergraduate', 'For-profit', 2, 0.335737,
          'Part-time', 'Undergraduate', 'For-profit', 3, 0.335737,
          'Part-time', 'Undergraduate', 'For-profit', -3, 0.335737,
          'Part-time', 'Graduate', 'For-profit', 1, 0.382059,
          'Part-time', 'Graduate', 'For-profit', 2, 0.335737,
          'Part-time', 'Graduate', 'For-profit', 3, 0.335737,
          'Part-time', 'Graduate', 'For-profit', -3, 0.335737)

# open database connection
db <- dbConnect(RSQLite::SQLite(), str_c(getwd(), 'data/db.sqlite', sep='/'))
```

# Read Data

## IPEDS Institutions

```{r}
# access data via net
system.time({
  ic <-
   tbl(db, 'characteristics') %>%
   filter(distnced == 0,
          ft_ug == 1) %>%
   select(unitid, year_key)

   institutions <- 
   tbl(db, 'directory') %>%
   filter(level_institution == 1,
          bea_region < 9,
          title4_eligible == 1,
          degree_granting == 1,
          control %in% 1:2)
    
 institutions <-
   institutions %>%
   inner_join(ic, by = c('unitid', 'year_key')) %>%
   group_by(unitid) %>%
   summarize(year_key = max(year_key)) %>%
   ungroup() %>%
   inner_join(institutions, by = c('unitid', 'year_key')) %>%
   mutate(close_year = ifelse(close_year > 0, close_year, NA),
          Closed = ifelse(is.na(close_year), 0, 1)) %>%
   select(unitid,
          `Institution Name` = institution_name,
          City = city,
          State = state,
          FIPS = fips,
          Level = level_institution,
          Control = control,
          Closed,
          `Close Year` = close_year,
          `Close Date` = close_date,
          Longitude = longitude,
          Latitude = latitude) %>%
   collect() %>%
   mutate(Control = recode(Control,
                           `1` = 'Public',
                           `2` = 'Private',
                           .default = 'Unknown'),
          `Close Date` = as_date(`Close Date`, tz = 'GMT', format='%m/%d/%Y'))

  # housekeeping
  rm(ic)
})
          
```

## Fall Enrollment

```{r}
system.time({
  # flags to roll up enrollments to match finance
  submissions <-
    tbl(db, 'submissions') %>%
    filter(year_key >= year_param) %>%
    mutate(parent_id = ifelse(prch_f %in% c('1', '2', '3'), idx_f, unitid)) %>%
    select(unitid, year_key, parent_id)
  
  # total_enrollment = fall enrollment (EF) table
  total_enrollment <-
    tbl(db, 'fall_enrollment') %>%
    inner_join(submissions, by = c('unitid', 'year_key')) %>%
    group_by(parent_id, year_key, time_status, career_level) %>%
    summarize(headcount = sum(headcount, na.rm = TRUE)) %>%
    rename(unitid = parent_id) %>%
    collect()
})
```

## IPEDS F2 Data Files

```{r}
# access data via net
system.time({
  gdp <-
    tbl(db, 'gdp') %>%
    filter(year_key >= year_param,
           month == 4) %>%
    select(year_key, gdp)

    finance <- 
      tbl(db, 'finance') %>%
      inner_join(gdp, by = 'year_key') %>%
      filter(field %in% c('f2b01', 'f2b02', 'f2d01', 'f2a03a', 'f2a17',
                          'f2a18', 'f2c05', 'f2c06', 'f2c08', 'f2c10',
                          'f2d01', 'f2e135', 'f2h02',
                          'f1d01', 'f1d02', 'f1a07', 'f1a27t4',
                          'f1a284', 'f1e05', 'f1e06', 'f1e08', 'f1e10',
                          'f1b01', 'f1c19dp', 'f1h02')) %>%
      collect() %>%
      mutate(amount = ifelse(is.na(amount), 0, amount / gdp),
             field = recode(field,
                            'f2b01' = 'Total Revenues',
                            'f2b02' = 'Total Expenses',
                            'f2a03a' = 'Total Debt',
                            'f2a17' = 'Total Plant',
                            'f2a18' = 'Accumulated Depreciation',
                            'f2c05' = 'Institutional Grants (Funded)',
                            'f2c06' = 'Institutional Grants (Unfunded)',
                            'f2c08' = 'Allowances to Tuition and Fees',
                            'f2c10' = 'Total Discounts',
                            'f2d01' = 'Tuition Revenue',
                            'f2e135' = 'Depreciation Expense',
                            'f2h02' = 'Endowment EOY',
                            'f1d01' = 'Total Revenues',
                            'f1d02' = 'Total Expenses',
                            'f1a07' = 'Total Debt',
                            'f1a27t4' = 'Total Plant',
                            'f1a284' = 'Accumulated Depreciation',
                            'f1e05' = 'Institutional Grants (Funded)',
                            'f1e06' = 'Institutional Grants (Unfunded)',
                            'f1e08' = 'Allowances to Tuition and Fees',
                            'f1e10' = 'Total Discounts',
                            'f1b01' = 'Tuition Revenue',
                            'f1c19dp' = 'Depreciation Expense',
                            'f1h02' = 'Endowment EOY')) %>%
      select(-gdp) %>%
      pivot_wider(names_from = field,
                  values_from = amount)
    
    rm(gdp)
})
```

## Get Year Labels

```{r}
# access data via net
system.time({
  years <- 
    tbl(db, 'ipeds_years') %>%
    select(year_key,
           fiscal_year,
           academic_year,
           calendar_year_fall) %>%
    collect()
})

```

```{r}
# Disconnect database
dbDisconnect(db)
  
# housekeeping
rm(submissions, db)
```

## Estimate FTE from Fall Enrollment Counts

```{r}
fte <-
  total_enrollment %>%
  inner_join(institutions, by = c('unitid')) %>%
  inner_join(weights, by = c('time_status', 
                             'career_level',
                             'Control',
                             'Level')) %>%
  mutate(fte = headcount * weight) %>%
  group_by(unitid, year_key, career_level) %>%
  summarize(Headcount = sum(headcount, na.rm = TRUE),
            FTE = sum(fte, na.rm = TRUE)) %>%
  ungroup() %>%
  pivot_wider(names_from = career_level,
              values_from = c(Headcount, FTE)) %>%
  replace_na(replace = list(Headcount_Graduate = 0,
                            Headcount_Undergraduate = 0,
                            FTE_Graduate = 0,
                            FTE_Undergraduate = 0)) %>%
  mutate(`Total FTE` = FTE_Undergraduate + FTE_Graduate,
         `Total Heacount` = Headcount_Undergraduate + Headcount_Graduate)

rm(total_enrollment, weights)
```

# Merge Files

```{r}
metrics <-
  institutions %>%
  inner_join(fte, by = c('unitid')) %>%
  inner_join(finance, by = c('unitid', 'year_key')) %>%
  inner_join(years, by = 'year_key') %>%
  select(-year_key) %>%
  rename(Unitid = unitid,
         `Fiscal Year` = fiscal_year,
         `Academic Year` = academic_year,
         `Calendar Year (Fall)` = calendar_year_fall) %>%
  mutate(Margin = `Total Revenues` - `Total Expenses`,
         `Margin per FTE (000)` = Margin / 1000 / `Total FTE`,
         `Gross tuition` = `Tuition Revenue` + `Allowances to Tuition and Fees`,
         `Tuition dependence` = `Tuition Revenue` / `Total Revenues`,
         `Discount rate` = `Allowances to Tuition and Fees` / `Gross tuition`,
         `Percentage of Grants from Unfunded Sources` = `Institutional Grants (Unfunded)` / `Total Discounts`,
         `Age of Plant` = `Accumulated Depreciation` / `Depreciation Expense`)

metrics %>%
  write_csv('data/ipeds_metrics.csv',
            na = '')
```

# References

```{r}
citation('base')
citation('broom')
citation('DBI')
citation('dbplyr')
citation('knitr')
citation('lubridate')
citation('odbc')
citation('RSQLite')
citation('tidyverse')
```

Built with R version `r getRversion()`.

Packages and versions:

broom `r packageVersion('broom')`.

DBI `r packageVersion('DBI')`.

dbplyr `r packageVersion('dbplyr')`.

dplyr `r packageVersion('dplyr')`.

ggplot `r packageVersion('ggplot2')`.

lubridate `r packageVersion('lubridate')`.

odbc `r packageVersion('odbc')`.

purrr `r packageVersion('purrr')`.

RSQLite `r packageVersion('RSQLite')`.

tidyr `r packageVersion('tidyr')`.

tidyverse `r packageVersion('tidyverse')`.
