---
title: "IPEDS Financial Metrics"
subtitle: "NCES Data Files"
author: "Jason P. Casey"
date: 2020-04-03
output: html_notebook
params:
  year:
    value: 2011
---

```{r setup}
# libraries
library(knitr)
library(DBI)
library(odbc)
library(dbplyr)
library(RSQLite)
library(lubridate)
library(broom)
library(tidyverse)

# set default document options
knitr::opts_chunk$set(fig.width = 6,
                      fig.asp = 0.618,
                      fig.retina = 2,
                      out.width = "70%",
                      fig.align = "center",
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      collapse = TRUE)

# set random seed
set.seed(1965)

# capture year_param for queries
year_param <- params$year

# EF headcount FTE weights
# see 'Calculation of FTE students (using fall student heacounts) in
#    https://surveys.nces.ed.gov/ipeds/VisGlossaryAll.aspx
weights <-
  tribble(~time_status, ~career_level, ~Control, ~Level, ~weight,
          # full-time weights are all 1.0
          'Full-time', 'Undergraduate', 'Public', 1, 1.0,
          'Full-time', 'Undergraduate', 'Public', 2, 1.0,
          'Full-time', 'Undergraduate', 'Public', 3, 1.0,
          'Full-time', 'Undergraduate', 'Public', -3, 1.0,
          'Full-time', 'Graduate', 'Public', 1, 1.0,
          'Full-time', 'Graduate', 'Public', 2, 1.0,
          'Full-time', 'Graduate', 'Public', 3, 1.0,
          'Full-time', 'Graduate', 'Public', -3, 1.0,
          'Full-time', 'Undergraduate', 'Private', 1, 1.0,
          'Full-time', 'Undergraduate', 'Private', 2, 1.0,
          'Full-time', 'Undergraduate', 'Private', 3, 1.0,
          'Full-time', 'Undergraduate', 'Private', -3, 1.0,
          'Full-time', 'Graduate', 'Private', 1, 1.0,
          'Full-time', 'Graduate', 'Private', 2, 1.0,
          'Full-time', 'Graduate', 'Private', 3, 1.0,
          'Full-time', 'Graduate', 'Private', -3, 1.0,
          'Full-time', 'Undergraduate', 'For-profit', 1, 1.0,
          'Full-time', 'Undergraduate', 'For-profit', 2, 1.0,
          'Full-time', 'Undergraduate', 'For-profit', 3, 1.0,
          'Full-time', 'Undergraduate', 'For-profit', -3, 1.0,
          'Full-time', 'Graduate', 'For-profit', 1, 1.0,
          'Full-time', 'Graduate', 'For-profit', 2, 1.0,
          'Full-time', 'Graduate', 'For-profit', 3, 1.0,
          'Full-time', 'Graduate', 'For-profit', -3, 1.0,
          # part-time
          'Part-time', 'Undergraduate', 'Public', 1, 0.403543,
          'Part-time', 'Undergraduate', 'Public', 2, 0.335737,
          'Part-time', 'Undergraduate', 'Public', 3, 0.335737,
          'Part-time', 'Undergraduate', 'Public', -3, 0.382059,
          'Part-time', 'Graduate', 'Public', 1, 0.361702,
          'Part-time', 'Graduate', 'Public', 2, 0.335737,
          'Part-time', 'Graduate', 'Public', 3, 0.335737,
          'Part-time', 'Graduate', 'Public', -3, 0.335737,
          'Part-time', 'Undergraduate', 'Private', 1, 0.392857,
          'Part-time', 'Undergraduate', 'Private', 2, 0.335737,
          'Part-time', 'Undergraduate', 'Private', 3, 0.335737,
          'Part-time', 'Undergraduate', 'Private', -3, 0.335737,
          'Part-time', 'Graduate', 'Private', 1, 0.382059,
          'Part-time', 'Graduate', 'Private', 2, 0.335737,
          'Part-time', 'Graduate', 'Private', 3, 0.335737,
          'Part-time', 'Graduate', 'Private', -3, 0.335737,
          'Part-time', 'Undergraduate', 'For-profit', 1, 0.392857,
          'Part-time', 'Undergraduate', 'For-profit', 2, 0.335737,
          'Part-time', 'Undergraduate', 'For-profit', 3, 0.335737,
          'Part-time', 'Undergraduate', 'For-profit', -3, 0.335737,
          'Part-time', 'Graduate', 'For-profit', 1, 0.382059,
          'Part-time', 'Graduate', 'For-profit', 2, 0.335737,
          'Part-time', 'Graduate', 'For-profit', 3, 0.335737,
          'Part-time', 'Graduate', 'For-profit', -3, 0.335737)
```

# Read Data

## Connect to Database

```{r}
# open database connection
db <- dbConnect(RSQLite::SQLite(), paste(getwd(), 'data/db.sqlite', sep = '/'))
```

## IPEDS Data

### Institutions

```{r}
# access data via net
system.time({
  ic <-
   tbl(db, 'characteristics') |>
   filter(distnced == 0,
          ft_ug == 1) |>
   select(unitid, year_key)

   institutions <- 
   tbl(db, 'directory') |>
   filter(level_institution == 1,
          bea_region < 9,
          title4_eligible == 1,
          degree_granting == 1,
          control %in% 1:2)
    
 institutions <-
   institutions |>
   inner_join(ic, by = c('unitid', 'year_key')) |>
   group_by(unitid) |>
   summarize(year_key = max(year_key)) |>
   ungroup() |>
   inner_join(institutions, by = c('unitid', 'year_key')) |>
   mutate(close_year = ifelse(close_year > 0, close_year, NA),
          Closed = ifelse(is.na(close_year), 0, 1)) |>
   select(unitid,
          `Institution Name` = institution_name,
          City = city,
          State = state,
          FIPS = fips,
          Level = level_institution,
          Control = control,
          Closed,
          `Close Year` = close_year,
          `Close Date` = close_date,
          Longitude = longitude,
          Latitude = latitude) |>
   collect() |>
   mutate(Control = recode(Control,
                           `1` = 'Public',
                           `2` = 'Private',
                           .default = 'Unknown'),
          `Close Date` = as_date(`Close Date`, tz = 'GMT', format='%m/%d/%Y'))

  # housekeeping
  rm(ic)
})
          
```

### Fall Enrollment

```{r}
system.time({
  # flags to roll up enrollments to match finance
  submissions <-
    tbl(db, 'submissions') |>
    filter(year_key >= year_param) |>
    mutate(parent_id = ifelse(prch_f %in% c('1', '2', '3'), idx_f, unitid)) |>
    select(unitid, year_key, parent_id)
  
  # total_enrollment = fall enrollment (EF) table
  total_enrollment <-
    tbl(db, 'fall_enrollment') |>
    inner_join(submissions, by = c('unitid', 'year_key')) |>
    group_by(parent_id, year_key, time_status, career_level) |>
    summarize(headcount = sum(headcount, na.rm = TRUE)) |>
    rename(unitid = parent_id) |>
    collect() |>
    ungroup()
})
```

### F1A and F2 Financial Data

```{r}
# access data via net
system.time({
  gdp <-
    tbl(db, 'gdp') |>
    filter(year_key >= year_param,
           month == 4) |>
    select(year_key, gdp)

    finance <- 
      tbl(db, 'finance') |>
      inner_join(gdp, by = 'year_key') |>
      filter(field %in% c('f2b01', 'f2b02', 'f2d01', 'f2a03a', 'f2a17',
                          'f2a18', 'f2c05', 'f2c06', 'f2c08', 'f2c10',
                          'f2d01', 'f2e135', 'f2h02',
                          'f1d01', 'f1d02', 'f1a07', 'f1a27t4',
                          'f1a284', 'f1e05', 'f1e06', 'f1e08', 'f1e10',
                          'f1b01', 'f1c19dp', 'f1h02')) |>
      collect() |>
      mutate(amount = ifelse(is.na(amount), 0, amount / gdp),
             field = recode(field,
                            'f2b01' = 'Total Revenues',
                            'f2b02' = 'Total Expenses',
                            'f2a03a' = 'Total Debt',
                            'f2a17' = 'Total Plant',
                            'f2a18' = 'Accumulated Depreciation',
                            'f2c05' = 'Institutional Grants (Funded)',
                            'f2c06' = 'Institutional Grants (Unfunded)',
                            'f2c08' = 'Allowances to Tuition and Fees',
                            'f2c10' = 'Total Discounts',
                            'f2d01' = 'Tuition Revenue',
                            'f2e135' = 'Depreciation Expense',
                            'f2h02' = 'Endowment EOY',
                            'f1d01' = 'Total Revenues',
                            'f1d02' = 'Total Expenses',
                            'f1a07' = 'Total Debt',
                            'f1a27t4' = 'Total Plant',
                            'f1a284' = 'Accumulated Depreciation',
                            'f1e05' = 'Institutional Grants (Funded)',
                            'f1e06' = 'Institutional Grants (Unfunded)',
                            'f1e08' = 'Allowances to Tuition and Fees',
                            'f1e10' = 'Total Discounts',
                            'f1b01' = 'Tuition Revenue',
                            'f1c19dp' = 'Depreciation Expense',
                            'f1h02' = 'Endowment EOY')) |>
      select(-gdp) |>
      pivot_wider(names_from = field,
                  values_from = amount)
    
    rm(gdp)
})
```

## Year Labels

Assigns fiscal and calendar years, aligning the collection cycles for IPEDS Data.

```{r}
# access data via net
system.time({
  years <- 
    tbl(db, 'ipeds_years') |>
    select(year_key,
           fiscal_year,
           academic_year,
           calendar_year_fall) |>
    collect()
})

```

## Disconnect from Database

```{r}
# Disconnect database
dbDisconnect(db)
  
# housekeeping
rm(submissions, db)
```

## Estimate FTE from Fall Enrollment Counts

```{r}
system.time({
  fte <-
    total_enrollment |>
    inner_join(institutions, by = c('unitid')) |>
    inner_join(weights, by = c('time_status', 
                               'career_level',
                               'Control',
                               'Level')) |>
    mutate(fte = headcount * weight) |>
    group_by(unitid, year_key, career_level) |>
    summarize(Headcount = sum(headcount, na.rm = TRUE),
              FTE = sum(fte, na.rm = TRUE)) |>
    ungroup() |>
    pivot_wider(names_from = career_level,
                values_from = c(Headcount, FTE)) |>
    replace_na(replace = list(Headcount_Graduate = 0,
                              Headcount_Undergraduate = 0,
                              FTE_Graduate = 0,
                              FTE_Undergraduate = 0)) |>
    mutate(`Total FTE` = FTE_Undergraduate + FTE_Graduate,
           `Total Heacount` = Headcount_Undergraduate + Headcount_Graduate)
  
  rm(total_enrollment, weights)
})
```

# Merge Files and Write Data File

```{r}
system.time({
  metrics <-
    institutions |>
    inner_join(fte, by = c('unitid')) |>
    inner_join(finance, by = c('unitid', 'year_key')) |>
    inner_join(years, by = 'year_key') |>
    select(-year_key) |>
    rename(Unitid = unitid,
           `Fiscal Year` = fiscal_year,
           `Academic Year` = academic_year,
           `Calendar Year (Fall)` = calendar_year_fall) |>
    mutate(Margin = `Total Revenues` - `Total Expenses`,
           `Margin per FTE (000)` = Margin / 1000 / `Total FTE`,
           `Gross tuition` = `Tuition Revenue` + `Allowances to Tuition and Fees`,
           `Tuition dependence` = `Tuition Revenue` / `Total Revenues`,
           `Discount rate` = `Allowances to Tuition and Fees` / `Gross tuition`,
           `Percentage of Grants from Unfunded Sources` = `Institutional Grants (Unfunded)` / `Total Discounts`,
           `Age of Plant` = `Accumulated Depreciation` / `Depreciation Expense`)
  
  metrics |>
    write_csv('data/ipeds_metrics.csv',
              na = '')
})
```

# Plotting Examples

## Bar Plot

```{r}
metrics |>
  filter(`Fiscal Year` == '2017-2018',
         Unitid %in% c('181464',
                       '181215',
                       '181020',
                       '181002',
                       '181394',
                       '181783',
                       '181446',
                       '181738',
                       '181428')) |>
  mutate(`Institution Name` = fct_reorder(`Institution Name`, `Tuition dependence`)) |>
  ggplot(aes(x = `Tuition dependence`,
             y = `Institution Name`,
             fill = Control)) +
    geom_col(alpha = 0.8) +
    scale_x_continuous(labels = scales::label_percent(),
                       limits = c(0,1)) +
    scale_fill_manual(values = c('darkgray', '#fc8d59')) +
    labs(title = 'Tuition Dependence at Selected Institutions',
         subtitle = 'FY2017-18',
         y = NULL)

```

## Scatter Plot of Tuition Revenue and Tuition Dependence

```{r}
metrics |>
  filter(`Fiscal Year` == '2017-2018',
         Control %in% c('Private', 'Public'),
         !is.na(Margin)) |>
  mutate(`Margin Category` = ifelse(Margin < 0, 'Loss', 'Even'),
         `Margin Category` = ifelse(Margin > 0, 'Gain', `Margin Category`),
         `Institutional Grants (Unfunded)` = `Institutional Grants (Unfunded)` / `Total FTE`) |>
  ggplot(aes(x = `Tuition dependence`,
             y = `Institutional Grants (Unfunded)`,
             color = `Margin Category`)) +
    geom_point(alpha = 0.5) +
    scale_x_continuous(labels = scales::label_percent(),
                       limits = c(0, 1)) +
    scale_y_continuous(breaks = scales::breaks_extended(8),
                       labels = scales::label_dollar(scale = .001,
                                                     suffix = 'K')) +
    scale_color_manual(values = c('#636363', '#91bfdb', '#fc8d59')) +
    labs(title = 'Average Unfunded Grant vs. Tuition Dependence',
         subtitle = 'Public and Private U.S. Institutions, FY2017-18')
```

## Heatmap

```{r}
metrics |>
  filter(`Fiscal Year` == '2017-2018',
         State == 'NE') |>
  mutate(`Revenues per FTE` = `Total Revenues` / `Total FTE`,
         `Tuition per FTE` = `Tuition Revenue` / `Total FTE`,
         `Discount per FTE` = `Total Discounts` / `Total FTE`,
         `Discount per FTE` = max(`Discount per FTE`) - `Discount per FTE`,
         `Average Unfunded Grant` = `Institutional Grants (Unfunded)` / `Total FTE`,
         `Endowment per FTE` = `Endowment EOY` / `Total FTE`,
         `Tuition dependence` = 1 - `Tuition dependence`) |>
  select(`Institution Name`,
         Control,
         `Revenues per FTE`,
         `Tuition per FTE`,
         `Discount per FTE`,
         `Average Unfunded Grant`,
         `Endowment per FTE`,
         `Tuition dependence`
         ) |>
  drop_na() |>
  pivot_longer(cols = 3:8,
               names_to = 'variable') |>
  group_by(variable) |>
  mutate(rescale = scale(value)) |>
  ungroup() |>
  ggplot(aes(x = variable,
             y = `Institution Name`)) +
    geom_tile(aes(fill = rescale),
              color = 'white') +
    scale_fill_gradient(low = '#ef8a62',
                        high = 'white') +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0)) +
    theme(legend.position = 'none',
          axis.text.x = element_text(angle = 45,
                                     hjust = 1)
          ) +
    labs(title = 'Selected Indicators',
         subtitle = 'Nebraska Institutions, FY2017-18',
         y = NULL,
         x = NULL)

```

# References

```{r}
citation('base')
citation('broom')
citation('DBI')
citation('dbplyr')
citation('knitr')
citation('lubridate')
citation('odbc')
citation('RSQLite')
citation('tidyverse')
```

Built with R version `r getRversion()`.

Packages and versions:

broom `r packageVersion('broom')`.

DBI `r packageVersion('DBI')`.

dbplyr `r packageVersion('dbplyr')`.

dplyr `r packageVersion('dplyr')`.

ggplot `r packageVersion('ggplot2')`.

lubridate `r packageVersion('lubridate')`.

odbc `r packageVersion('odbc')`.

purrr `r packageVersion('purrr')`.

RSQLite `r packageVersion('RSQLite')`.

tidyr `r packageVersion('tidyr')`.

tidyverse `r packageVersion('tidyverse')`.
