---
title: "IPEDS Financial Metrics"
subtitle: "Visualizations"
author: "Jason P. Casey"
date: 2020-06-14
output: 
  html_notebook:
    code_folding: show
params:
  unitid:
    label: "Reference Institution"
    value: 181464
---
                      
```{r setup}
# libraries
library(knitr)
library(janitor)
library(lubridate)
library(broom)
library(tidyverse)
library(data.table)
library(RColorBrewer)

# set default document options
knitr::opts_chunk$set(fig.width = 6,
                      fig.asp = 0.618,
                      fig.retina = 2,
                      out.width = "70%",
                      fig.align = "center",
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      collapse = TRUE)

# set random seed
set.seed(1965)
```

# Read Data

```{r}
system.time({
  # read stress scores file
  stress <-
    fread('data/stress.csv')
  
  # cutoffs
  criteria <- data.table(
    Metric = c('UG First-time First Year Enrollment',
               'Retention',
               'State Appropriations',
               'Market Price',
               'UG First-time First Year Enrollment',
               'Retention',
               'Endowment/Expenses',
               'Market Price'),
    Control = c(rep('Public', 5), rep('Private', 5)),
    Alert = c(-0.154, 0.68, -0.27, 0, -0.174, 0.656, 0, -0.104),
    Warning = c(-0.256, 0.62, -0.37, -0.083, -0.308, 0.564, -0.103, -0.175)
  )
  
  metrics <-
    melt.data.table(
      fread('data/metrics.csv')[order(Unitid, year_key)][,
                                                         .(Unitid,
                                                          `Institution Name`,
                                                          State,
                                                          Control,
                                                          year_key,
                                                          `Fiscal Year`,
                                                          `UG First-time First Year Enrollment`,
                                                          Retention,
                                                          `Market Price`,
                                                          `Endowment/Expenses`,
                                                          `State Appropriations`)],
      measure.vars = 7:11,
      variable.name = 'Metric',
      value.name = 'Value'
    ) %>%
    # join criteria
    inner_join(criteria, by = c('Control', 'Metric')) %>%
    # add normed values and convert alerts and warnings to proper scale
    .[,
      `:=`(`Normed Value` = Value / first(Value) * 100,
           Alert = ifelse(Metric == 'Retention', Alert, first(Value) * (1 + Alert)),
           Warning = ifelse(Metric == 'Retention', Warning, first(Value) * (1 + Warning))),
      keyby = .(Unitid, Metric)]

    # get reference institution and sector
  inst <- stress[Unitid == params$unitid, `Institution Name`, drop = TRUE]
  sector <- stress[Unitid == params$unitid, Control, drop = TRUE]
  
  # metric names for sector of institution
  metric_names <- ifelse(sector == 'Public', 'State Appropriations', 'Endowment/Expenses') %>%
    c(., c('UG First-time First Year Enrollment', 'Retention', 'Market Price'))


  # housekeeping
  rm(criteria)
})
```

# Reference Institution Metric Values

```{r}
metrics[Unitid == params$unitid]
```

# Reference Institution Stess Score

```{r}
stress[Unitid == params$unitid]
```

# Plots of Reference Institution Versus Sector

## First Year Enrollment

```{r}
# function to create standard-formatted box plots with reference institution
box_me <- function(metric)
{
  df <- metrics[Control == sector & Metric == metric]
  
  ggplot(data = df,
         aes(x = `Fiscal Year`, y = Value)) +
    # box plot for reference group
    geom_boxplot(color = '#1f78b4', fill='#a6cee3', alpha = 1/5) +
    # line plot for reference institution
    geom_line(data = df[Unitid == params$unitid],
              aes(x = `Fiscal Year`, y = Value, group = 1),
              color = 'black',
              size = 1.5) +
    # points for reference institution
    geom_point(data = df[Unitid == params$unitid],
               aes(x = `Fiscal Year`, y = Value),
               color = 'black',
               shape = 18,
               size = 6) +
    labs(title = metric,
         y = metric,
         subtitle = str_c(inst, 'vs. Four Year', sector, 'Institutions', sep = ' '))
}



box_me('UG First-time First Year Enrollment') +
  # set y-axis to comma format
  scale_y_continuous(labels = scales::comma_format()) +
  labs(title = 'First Year Undergraduate Enrollment')
```

## Retention

```{r}
box_me('Retention') +
  # set y-axis to percentage format
  scale_y_continuous(labels = scales::label_percent())
```

## Market Price

```{r}
box_me('Market Price') +
  # set y-axis to dollar format
  scale_y_continuous(labels = scales::label_dollar()) +
  labs(caption = 'Dollars adjusted for inflation using GDP.')
```

## Endowment/Expenses (Privates) or State Appropriations (Publics)

```{r}
if (sector == 'Public') {
  box_me('State Appropriations') +
    # set y-axis to dollar format
    scale_y_continuous(labels = scales::label_dollar(scale = .000001,
                                                     suffix = 'M'))
} else {
  box_me('Endowment/Expenses') +
    # set y-axis to dollar format
    scale_y_continuous(labels = scales::label_dollar()) +
    labs(title = 'Endowment to Total Expenses less Hospital Expenses')
}
```

# Plot of Normed Metrics


```{r}
ggplot(data = metrics[Unitid == params$unitid],
       mapping = aes(x = `Fiscal Year`,
                     y = `Normed Value`,
                     group = Metric,
                     color = Metric,
                     linetype = Metric)) +
  geom_line(size = 1.5) +
  labs(title = str_c('Scaled Metrics for', inst, sep = ' '),
       subtitle = str_c(min(metrics$`Fiscal Year`), max(metrics$`Fiscal Year`), sep = ' through '),
       caption = 'Scaled to First Year = 100',
       y = 'Normed Value',
       color = 'Metric') +
  scale_colour_brewer(palette = 'Set2')
```

# Plots of Institutional Metrics

## First Year Enrollment

```{r}
# statistical model specifications
run_model <- function(df) {
  lm(formula = Value ~ year_key, 
     data = df)
}

plot_me <-
  function(metric) {
    labels <-
      metrics %>%
      filter(Unitid == params$unitid,
             Metric == metric) %>%
      distinct(Alert, Warning) %>%
      pivot_longer(1:2, names_to = 'label', values_to = 'Value') %>%
      mutate(`Fiscal Year` = '2010-2011')
    
    metrics %>%
      select(Unitid, `Institution Name`, State, Control, Metric, Alert, Warning) %>%
      expand_grid(year_key = seq(max(metrics$year_key) + 1, max(metrics$year_key) + 3)) %>%
      mutate(`Fiscal Year` = str_c(year_key - 1, year_key, sep='-')) %>%
      bind_rows(metrics, .) %>%
      filter(Unitid == params$unitid) %>%
      group_by(Unitid, Metric) %>%
      nest() %>%
      ungroup() %>%
      mutate(fit = map(data, run_model),
             Trend = map2(fit, data, predict)) %>%
      select(-fit) %>%
      unnest(cols = c('data', 'Trend')) %>%
      filter(Metric == metric) %>%
      ggplot(aes(x = `Fiscal Year`,
                 y = Value,
                 group = 1)) +
      geom_line(size = 1.5) +
    # add linear trend line
    geom_line(aes(y = Trend),
              linetype = 'longdash',
              size = 1,
              color = '#7570b3') +
    # add alert and warning lines
    geom_hline(aes(yintercept = Alert),
               linetype = 'dotted',
               color = '#1b9e77',
               size = 1) +
    geom_hline(aes(yintercept = Warning),
               linetype = 'dotted',
               color = '#d95f02',
               size = 1) +
    # Add line labels
    geom_text(aes(label = label,
                  y = Value),
              data = labels,
              hjust = "left") +
    labs(title = metric,
         subtitle = inst,
         y = metric)
  }

plot_me('UG First-time First Year Enrollment') +
  # set y-axis to comma format
  scale_y_continuous(labels = scales::label_comma()) +
  labs(title = 'Undergraduate First Year Enrollment')
```

## Retention

```{r}
plot_me('Retention') +
  # set y-axis to comma format
  scale_y_continuous(labels = scales::label_percent())
```

## Market Price

```{r}
plot_me('Market Price') +
  # set y-axis to comma format
  scale_y_continuous(labels = scales::label_dollar())
```

## Endowment/Expenses (Privates) or State Appropriations (Publics)

```{r}
if (sector == 'Public') {
  plot_me('State Appropriations') +
    # set y-axis to dollar format
    scale_y_continuous(labels = scales::label_dollar(scale = .000001,
                                                     suffix = 'M'))
} else {
  plot_me('Endowment/Expenses') +
    # set y-axis to comma format
    scale_y_continuous(labels = scales::label_comma(scale = 2)) +
    labs(title = 'Endowment to Expenses',
         caption = 'Expenses = Total Expenses net of Hospital Expenses')
}
```

# Other Useful Stuff

## Color Palette

Some people have difficulty with certain color palettes, so I've used colors from one of the colorblind "safe" palettes in the default plots.  However, these may not work for all users -- they may have different sensitivities or may prefer a different choice of colors.  If you prefer a different palette and want to use a default, the resources below provide some guidance on using the [Color Brewer](https://colorbrewer2.org/#type=qualitative&scheme=Accent&n=3) palettes.

### Palette Names

```{r}
brewer.pal.info
```

### Color Brewer Qualitative (Categorical) Palettes

```{r}
display.brewer.all(n=NULL, type="qual", select=NULL, exact.n=TRUE, 
colorblindFriendly=FALSE)
```
## Save Outfiles

The following will save the long format data to two files:

* data/normed_metrics.csv to be used by software that can read CSV files
* app/data/normed_metrics.RDS to be used by a Shiny app

```{r}
system.time({
  metrics %>%
    write_csv('data/normed_metrics.csv',
              na = '')
  
  metrics %>%
    saveRDS('app/data/normed_metrics.RDS')
})
```

