---
title: "IPEDS Financial Metrics"
subtitle: "NCES Data Files"
author: "Jason P. Casey"
date: 2020-04-05
output: 
  html_notebook:
    code_folding: show
params:
  year:
    label: "Last year of series"
    value: 2018
  unitid:
    label: "Reference Institution"
    value: "181464"
---

# Introduction

The following script estimates **College Stress Scores** as defined in _The College Stress Test: Tracking Institutional Futures across a Crowded Market_ by Robert Zemsky, Susan Shaman, and Susan Campbell Baldridge.  The IPEDS pieces are those recommended by the authors and estimations follow the guidelines in Appendix A of the book.

The authors limited their investigations to institutions that:
* have undergraduate enrollment
* are degree-granting
* are in the 50 states (and DC)
* are eligible to participate in Title IV financial aid

Additionally, this script limits the population to four-year institutions.  (Two year institutions may be examined separately).

Because an institution's IPEDS Finance is often reported as part of a parent institution's record, the parent unitid was needed from the FLAGS file to roll up the non-financial metrics to match the financials.

## Fields Used

File   | Field    | Description    | Notes
-------|----------|----------------|---------------------
HD     | unitid   | Unique identification number of the institution | 
HD     | instnm   | Institution (entity) name | 
HD     | city     | City location of institution |  
HD     | stabbr   | State abbreviation | 
HD     | fips     | FIPS state code | 
HD     | obereg   | Bureau of Economic Analysis (BEA) regions | _obereg < 9 _
HD     | opeflag  | OPE Title IV eligibility indicator code | _opeflag == 1_
HD     | control  | Control of institution | 
HD     | deggrant | Degree-granting status | _deggrant == 1_
HD     | longitud | Longitude location of institution | 
HD     | latitude | Latitude location of institution | 
HD     | iclevel  | Level of institution | _iclevel == 1_
IC     | ft_ug    | Full-time undergraduate students are enrolled | _ft_ug == 1_
IC     | distnced | All programs offered completely via distance education | _distnced == 0_
FLAGS  | prch_f   | Parent/child indicator- Finance | 
FLAGS  | idx_f    | ID number of parent institution - Finance | 
SFA    | igrnt_t  | Total amount of institutional grant aid awarded to full-time first-time undergraduates | _zero imputed for missing_
IC AY  | tuition2 | In-state average tuition for full-time undergraduates | 
IC AY  | tuition3 | Out-of-state average tuition for full-time undergraduates | 
EFA    | line     | Level of student (original line number on survey form) | 
EFA    | eftotlt  | Grand total | _(2008 and after)_
EFA    | efrace15 | Total men | _(prior to 2008)_
EFA    | efrace16 | Total women | _(prior to 2008)_
EFC    | line     | State of residence  (original line number on survey form) | _line not in (58, 99)_
EFC    | efres01  | First-time degree/certificate-seeking undergraduate students | _institutional mean imputed for missing odd years_
EFD    | ret_pcf  | Percent of first-time full-time degree/certificate-seeking undergraduate students in fall 2002 returning in fall 2003 | 
F1A    | f1b11    | State appropriations | 
F1A    | f1h02    | Value of endowment assets at the end of the fiscal year | 
F1A    | f1c191   | Total expenses-Total amount | 
F1A    | f1c121   | Hospital services-Total amount | 
F2     | f2d03    | State appropriations | 
F2     | f2h02    | Value of endowment assets at the end of the fiscal year | 
F2     | f2e131   | Total expenses-Total amount | 
F2     | f2e091   | Hospital services-Total amount | 

In addition to IPEDS, the model also uses Gross Domestic Product (GDP) deflators.  These are available quarterly from the Federal Researve Bank of St. Louis in csv format. Zemsky, et al, used the fall (October) quarter _following_ the end of the fiscal cycle for each year.  Each _gdp_ score is divided by 100 to make them proper defaltors.


```{r setup}
# libraries
library(knitr)
library(DBI)
library(odbc)
library(dbplyr)
library(RSQLite)
library(lubridate)
library(broom)
library(hexbin)
library(tidyverse)
library(data.table)

# set default document options
knitr::opts_chunk$set(fig.width = 6,
                      fig.asp = 0.618,
                      fig.retina = 2,
                      out.width = "70%",
                      fig.align = "center",
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      collapse = TRUE)

# set random seed
set.seed(1965)

# include user functions
source(file = "user_functions.R")
```

# Import Data

## Load Tables

```{r}
system.time({
  # create a file database (db) to store IPEDS data
  db <- dbConnect(RSQLite::SQLite(), paste(getwd(), 'data/db.sqlite', sep = '/'))
  
  ic <-
    tbl(db, 'characteristics') |>
    filter(distnced == 0,
           ft_ug == 1) |>
    select(unitid, year_key)
  
  year_param <- params$year
  
  years <-
    tbl(db, 'ipeds_years') |>
    filter(year_key > year_param - 8,
           year_key <= year_param) |>
    select(year_key, fiscal_year) |>
    as.data.table()
  
  years[, Index := row_number()]
  
  submissions <-
    tbl(db, 'submissions') |>
    mutate(parent_id = ifelse(prch_f %in% c('1', '2', '3'), idx_f, unitid)) |>
    select(unitid, year_key, parent_id) |>
    collect()
  
  institutions <- 
    tbl(db, 'directory') |>
    filter(level_institution == 1,
           bea_region < 9,
           bea_region > 0,
           title4_eligible == 1,
           degree_granting == 1)
    
  total_enrollment <-
    tbl(db, 'fall_enrollment') |>
    filter(year_key > year_param - 8,
           year_key <= year_param,
           career_level == 'Undergraduate',
           degree_seeking == 'Degree-seeking',
           continuation_type == 'First-time') |>
    group_by(unitid, year_key) |>
    summarize(`UG First-time First Year Enrollment` = sum(headcount, na.rm = TRUE)) |>
    collect() |>
    inner_join(submissions, by = c('unitid', 'year_key')) |>
    group_by(parent_id, year_key) |>
    summarize(`UG First-time First Year Enrollment` = sum(`UG First-time First Year Enrollment`, na.rm = TRUE)) |>
    rename(unitid = parent_id)
  
  enrollment_by_state <-
    tbl(db, 'enrollment_by_state') |>
    inner_join(institutions, by = c('unitid', 'year_key')) |>
    mutate(`In State Enrollment` = ifelse(fips.x == fips.y, headcount, 0),
           `Out of State Enrollment` = ifelse(!fips.x == fips.y, headcount, 0)) |>
    group_by(unitid, year_key) |>
    summarize(`In State Enrollment` = sum(`In State Enrollment`, na.rm = TRUE),
              `Out of State Enrollment` = sum(`Out of State Enrollment`, na.rm = TRUE)) |>
    collect() |>
    inner_join(submissions, by = c('unitid', 'year_key')) |>
    group_by(parent_id, year_key) |>
    summarize(`In State Enrollment` = sum(`In State Enrollment`, na.rm = TRUE),
              `Out of State Enrollment` = sum(`Out of State Enrollment`, na.rm = TRUE)) |>
    rename(unitid = parent_id)
  
  retention <-
    tbl(db, 'retention') |>
    filter(year_key > year_param - 8,
           year_key <= year_param,
           retention_ft > 0) |>
    select(unitid,
           year_key,
           Retention = retention_ft) |>
    collect() |>
    inner_join(submissions, by = c('unitid', 'year_key')) |>
    group_by(parent_id, year_key) |>
    summarize(Retention = mean(Retention, na.rm = TRUE)) |>
    rename(unitid = parent_id)
  
  academic_year_charges <-
    tbl(db, 'academic_year_charges') |>
    filter(year_key > year_param - 8,
           year_key <= year_param,
           field %in% c('tuition2', 'tuition3')) |>
    collect() |>
    pivot_wider(names_from = field,
                values_from = value) |>
    rename(`In State Tuition` = tuition2,
           `Out of State Tuition` = tuition3) |>
    inner_join(submissions, by = c('unitid', 'year_key')) |>
    group_by(parent_id, year_key) |>
    summarize(`In State Tuition` = mean(`In State Tuition`, na.rm = TRUE),
              `Out of State Tuition` = mean(`Out of State Tuition`, na.rm = TRUE)) |>
    rename(unitid = parent_id)
  
  student_financial_aid <-
    tbl(db, 'student_financial_aid') |>
    filter(year_key > year_param - 8,
           year_key <= year_param,
           field == 'igrnt_t') |>
    collect() |>
    pivot_wider(names_from = field,
                values_from = value) |>
    inner_join(submissions, by = c('unitid', 'year_key')) |>
    group_by(parent_id, year_key) |>
    summarize(igrnt_t = sum(igrnt_t, na.rm = TRUE)) |>
    rename(unitid = parent_id,
           `Inst Grant Aid` = igrnt_t)
    
  finance <-
    tbl(db, 'finance') |>
    filter(year_key > year_param - 8,
           year_key <= year_param,
           field %in% c('f1b11', 'f1h02', 'f1c191', 'f1c121',
                        'f2d03', 'f2h02', 'f2e131', 'f2e091')) |>
    collect() |>
    mutate(field = recode(field,
                          'f1b11' = 'State Appropriations',
                          'f1h02' = 'Endowment EOY',
                          'f1c191' = 'Total Expenses',
                          'f1c121' = 'Hospital Expenses',
                          'f2d03' = 'State Appropriations',
                          'f2h02' = 'Endowment EOY',
                          'f2e131' = 'Total Expenses',
                          'f2e091' = 'Hospital Expenses')) |>
    pivot_wider(names_from = field,
                values_from = amount)
    
  institutions <-
    institutions |>
    inner_join(ic, by = c('unitid', 'year_key')) |>
    group_by(unitid) |>
    summarize(year_key = max(year_key)) |>
    ungroup() |>
    inner_join(institutions, by = c('unitid', 'year_key')) |>
    mutate(close_year = ifelse(as.integer(close_year) > 0, close_year, NA)) |>
    select(unitid,
           `Institution Name` = institution_name,
           City = city,
           State = state,
           FIPS = fips,
           Control = control,
           `Close Year` = close_year,
           longitude,
           latitude) |>
    collect()
    
  gdp <-
    tbl(db, 'gdp') |>
    filter(year_key > year_param - 8,
           year_key <= year_param,
           month == 4) |>
    select(year_key, gdp) |>
    collect() |>
    inner_join(years, by = 'year_key')
  
  # Disconnect database
  dbDisconnect(db)
  rm(db)
})
```

## Merge Data Frames

```{r}
system.time({
  df <-
    institutions |>
    inner_join(submissions, by = 'unitid') |>
    inner_join(gdp, by = 'year_key') |>
    inner_join(retention, by = c('unitid', 'year_key')) |>
    inner_join(total_enrollment, by = c('unitid', 'year_key')) |>
    inner_join(student_financial_aid, by = c('unitid', 'year_key')) |>
    inner_join(academic_year_charges, by = c('unitid', 'year_key')) |>
    inner_join(finance, by = c('unitid', 'year_key')) |>
    left_join(enrollment_by_state, by = c('unitid', 'year_key')) |>
    add_count(unitid) |>
    filter(n == 8) |>
    select(-n) |>
    mutate(Control = recode(Control,
                            `1` = 'Public',
                            `2` = 'Private',
                            `3` = 'For-profit',
                            .default = 'Unknown')) |>
    rename(Unitid = unitid,
           `Accounting Standard` = accounting_standard,
           Longitude = longitude,
           Latitude = latitude,
           `Fiscal Year` = fiscal_year,
           GDP = gdp)
  
  # housekeeping
  rm(submissions,
     gdp,
     retention,
     total_enrollment,
     student_financial_aid,
     academic_year_charges,
     finance,
     enrollment_by_state,
     ic,
     ipeds_years,
     institutions)
})
```

## Apply Deflators

```{r}
system.time({
  adjusted <-
    df |>
    mutate(`Total Expenses` = `Total Expenses` / GDP,
           `Hospital Expenses` = `Hospital Expenses` / GDP,
           `Endowment EOY` = `Endowment EOY` / GDP,
           `Inst Grant Aid` = `Inst Grant Aid` / GDP,
           `In State Tuition` = `In State Tuition` / GDP,
           `Out of State Tuition` = `Out of State Tuition` / GDP,
           `State Appropriations` = `State Appropriations` / GDP)
})
```

## Imputations

```{r}
system.time({
  imputed <-
    adjusted |>
    group_by(Unitid) |>
    mutate(Retention = ifelse(is.na(Retention), median(Retention, na.rm = TRUE), Retention),
           `In State Enrollment` = ifelse(is.na(`In State Enrollment`),
                                          median(`In State Enrollment`,
                                                 na.rm = TRUE),
                                          `In State Enrollment`),
           `Out of State Enrollment` = ifelse(is.na(`Out of State Enrollment`),
                                              median(`Out of State Enrollment`,
                                                     na.rm = TRUE),
                                              `Out of State Enrollment`),
           `Total Expenses` = ifelse(is.na(`Total Expenses`),
                                     median(`Total Expenses`,
                                            na.rm = TRUE),
                                     `Total Expenses`),
           `Hospital Expenses` = ifelse(is.na(`Hospital Expenses`),
                                     median(`Hospital Expenses`,
                                            na.rm = TRUE),
                                     `Hospital Expenses`),
           `Endowment EOY` = ifelse(is.na(`Endowment EOY`),
                                     median(`Endowment EOY`,
                                            na.rm = TRUE),
                                     `Endowment EOY`),
           `Inst Grant Aid` = ifelse(is.na(`Inst Grant Aid`),
                                     median(`Inst Grant Aid`,
                                            na.rm = TRUE),
                                     `Inst Grant Aid`),
           `In State Tuition` = ifelse(is.na(`In State Tuition`),
                                     median(`In State Tuition`,
                                            na.rm = TRUE),
                                     `In State Tuition`),
           `Out of State Tuition` = ifelse(is.na(`Out of State Tuition`),
                                     median(`Out of State Tuition`,
                                            na.rm = TRUE),
                                     `Out of State Tuition`),
           `State Appropriations` = ifelse(is.na(`State Appropriations`),
                                     median(`State Appropriations`,
                                            na.rm = TRUE),
                                     `State Appropriations`)) |>
    ungroup()
})
```

## Create Base Metrics

Because EF-C is optional in odd-numbered years, there were a considerable number of missing values.  An institution's mean values were imputed for each missing year of counts for this section of the survey.  All non-financial metrics were aggregated at the parent institution's unitid and data were merged with the institutional header and financial data.  The basic metrics were computed as recommended in the book:

* _Undergraduate First-time First-year Enrollment_
* _Retention_ = Retention Rate
* _Endowment/Expenses_ = End-of-year Endowment / Total Expenses net of Hospital Expense
* _Market Price_ = In State Tuition times Proportion of In State Enrollment plus Out of State Tution time Proportion of Out of State Enrollment Less Institutional Grant Aid

```{r}
system.time({
  metrics <-
    imputed |>
    add_count(Unitid) |>
    filter(n == 8) |>
    mutate(`Total Enrollment` = `In State Enrollment` + `Out of State Enrollment`,
           `In State Enrollment Pct` = `In State Enrollment` / `Total Enrollment`,
           `Out of State Enrollment Pct` = `Out of State Enrollment` / `Total Enrollment`,
           `Endowment/Expenses` = `Endowment EOY` / (`Total Expenses` - `Hospital Expenses`),
           `Average Grant Aid` = `Inst Grant Aid` / `UG First-time First Year Enrollment`,
           `Market Price` = (`In State Tuition` * `In State Enrollment Pct` + `Out of State Tuition` * `Out of State Enrollment Pct`) - `Average Grant Aid`) |>
    select(-c(n, `Total Enrollment`))
})
```

## Output Metrics Files

```{r}
system.time({
  metrics |>
    write_csv('data/metrics.csv',
              na = '')
  saveRDS(metrics, 'data/metrics.rds')
})
```

## Inspect Institutions With Unusual Market Price

```{r}
metrics |>
  filter(`Market Price` <= 0)
```

## Estimate Institutional Extrapolation Models

The authors create a linear function for each institution and use this to project three years past the most recent year.

```{r}
# metrics <-
#   read_csv('data/metrics.csv')
# 
system.time({
  coef_helper <- function(td) as_tibble(t(deframe(td)))
  models <-
    metrics |>
    select(Unitid,
           Control,
           Index,
           `UG First-time First Year Enrollment`,
           Retention,
           `State Appropriations`,
           `Endowment/Expenses`,
           `Market Price`) |>
    drop_na() |>
    pivot_longer(cols = 4:8,
                 names_to = 'Metric',
                 values_to = 'Value') |>
    group_by(Unitid, Control, Metric) |>
    nest() |>
    mutate(model = map2('Value ~ Index', data, run_model),
           td = map(model, tidy),
           coef = map(td, coef_helper))
})
```

## Project Future Values

Using the estimated coefficients, project the value three years from the most recent year.

```{r}
system.time({
  projections <-
    models |>
    select(Unitid,
           Control,
           Metric,
           coef) |>
    unnest(cols = coef) |>
    rename(Slope = Index) |>
    mutate(Index = 11,
           Value = `(Intercept)` + Slope * Index) |>
    select(Unitid,
           Control,
           Index,
           Metric,
           Value) |>
    pivot_wider(names_from = Metric,
                values_from = Value)
})
```

## Compute Stress Scores

Change rates from the base year to the most recent year were calculated for four metrics for each institution type.

**Publics**

* Undergradaute First-time First-year Enrollment
* First-year to Second-year Retention Rate
* State Appropriations (in Constant Dollars)
* Market Price (in Constant Dollars)

**Privates**

* Undergradaute First-time First-year Enrollment
* First-year to Second-year Retention Rate
* Endowment/Expenses (in Constant Dollars)
* Market Price (in Constant Dollars)

These were compared to benchmarks established by the authors to set statuses of _Alert_ and _Warning_ (consistent with National Weather Service terminology) for each of the four core metrics.  These Alert and Warning Scores were summed to produce the final _Stress Score_.

_Notes_

* It is unclear whether the relationship between _Undergraduate First-time First-year Enrollment_ and _Market Price_ is fully accounted for in the authors' model.  Does a decrease in realized tuition reflect poor financial health if enrollments are increasing?  Similarly, increased tuition may not reflect strenth if enrollments are falling.
* Many institutions that have publicly acknowledged financial struggles do not seem to have correspondingly high stress scores.
* Missing values are hard to account for in the calculation of metrics and stress scores.  Non-random patterns of missing data might undermine this model for measuring risk.

```{r}
system.time({
  criteria <-
    tribble(~Metric, ~Control, ~Crit1, ~Crit2,
            'UG First-time First Year Enrollment', 'Public', -0.154, -0.256,
            'Retention', 'Public', 0.68, 0.62,
            'Endowment/Expenses', 'Public', 0, -0.103,
            'State Appropriations', 'Public', -0.27, -0.37,
            'Market Price', 'Public', 0, -0.083,
            'UG First-time First Year Enrollment', 'Private', -0.174, -0.308,
            'Retention', 'Private', 0.656, 0.564,
            'Endowment/Expenses', 'Private', 0, -0.103,
            'State Appropriations', 'Private', -0.27, -0.37,
            'Market Price', 'Private', -0.104,  -0.175)
  
  institution_header <-
    metrics |>
    distinct(Unitid,
             `Institution Name`,
             City,
             State,
             FIPS,
             Control,
             `Accounting Standard`,
             `Close Year`,
             Longitude,
             Latitude)
})
```

### Compute Change Scores

```{r}
system.time({
  change_scores <-
    metrics |>
    select(Unitid,
           Control,
           Index,
           `UG First-time First Year Enrollment`,
           Retention,
           `State Appropriations`,
           `Endowment/Expenses`,
           `Market Price`) |>
    bind_rows(projections) |>
    pivot_longer(cols = 4:8,
                 names_to = 'Metric',
                 values_to = 'Value') |>
    arrange(Unitid, Control, Metric, Index) |>
    group_by(Unitid, Control, Metric) |>
    mutate(Change = ifelse(Index == 8,
                           (Value - lag(Value, 7)) / lag(Value, 7),
                           (Value - lag(Value, 8)) / lag(Value, 8))
           ) |>
    ungroup() |>
    drop_na() |>
    mutate(Change = ifelse(Metric == 'Retention', Value, Change)) # Retention uses actual value, not change score
})
```

### Add Alert and Warning Scores

```{r}
system.time({
  warning_scores <-
    change_scores |>
    count(Unitid, Control, Metric) |>
    inner_join(change_scores) |>
    filter(n == 2) |>
    inner_join(criteria) |>
    mutate(Alert = ifelse(Change <= Crit1, 1, 0),
           Alert = ifelse(Metric == 'Retention' & Value <= Crit1, 1, Alert),
           Warning = ifelse(Change <= Crit2 & Index == 8, 1, 0),
           Warning = ifelse(Metric == 'Retention' & Index == 8 & Value <= Crit2, 1, Warning)) |>
    select(-n, -Value, -Crit1, -Crit2)
})

warning_scores
```

### Reshape Final Stress Scores

```{r}
system.time({
  stress_scores <-
    warning_scores |>
    pivot_longer(cols = 6:7,
                 names_to = 'Status',
                 values_to = 'Value') |>
    mutate(Value = ifelse(Control == 'Public' & Metric == 'Endowment/Expenses', 0, Value),
           Value = ifelse(Control == 'Private' & Metric == 'State Appropriations', 0, Value)) |>
    group_by(Unitid) |>
    summarize(`Stress Score` = sum(Value)) |>
    ungroup() |>
    mutate(`Stress Score Pct` = `Stress Score` / 12) |>
    inner_join(warning_scores, ., by = 'Unitid') |>
    mutate(Type = recode(Index,
                         `8` = 'Actual',
                         `11` = 'Projected')) |>
    select(-Index) |>
    pivot_wider(names_from = c(Metric, Type),
                values_from = c(Change, Alert, Warning)) |>
    inner_join(institution_header, ., by = c('Unitid', 'Control'))
})
```

## Output Stress Scores File

```{r}
system.time({
  stress_scores |>
    write_csv('data/stress.csv',
              na = '')
  saveRDS(stress_scores, 'data/stress.rds')
})
```

# Plots

## Endowment/Expenses vs Market Price

```{r}
metrics |>
  ggplot(aes(x = `Endowment/Expenses`, y = `Market Price`, color = Control)) +
    geom_point(alpha = 1/5) + # geom_hex(bins = 40) +
    facet_wrap(facets = vars(Control), nrow = 2, ncol = 1) +
    labs(title = 'Endowment-to-Expenses vs. Market Price',
         x = 'Endowment/Expenses',
         y = 'Market Price')
```

## State Appropriations vs Market Price

```{r}
metrics |>
  ggplot(aes(x = `State Appropriations`, y = `Market Price`, color = Control)) +
    geom_point(alpha = 1/5) + # geom_hex(bins = 40) +
    facet_wrap(facets = vars(Control), nrow = 2, ncol = 1) +
    labs(title = 'State Appropriations vs. Market Price',
         x = 'State Appropriations',
         y = 'Market Price')
```

# References

## Book

Zemsky, R., Shaman, S., Campbell Baldridge, S. (2020). _The College Stress Test: Tracking Institutional Futures across a Crowded Market_. Johns Hopkins University Press.  Baltimore, Maryland.

## Software

```{r}
citation('base')
citation('broom')
citation('DBI')
citation('dbplyr')
citation('hexbin')
citation('knitr')
citation('lubridate')
citation('odbc')
citation('RSQLite')
citation('tidyverse')
```

Built with R version `r getRversion()`.

Packages and versions:

broom `r packageVersion('broom')`.

DBI `r packageVersion('DBI')`.

dbplyr `r packageVersion('dbplyr')`.

dplyr `r packageVersion('dplyr')`.

ggplot `r packageVersion('ggplot2')`.

hexbin `r packageVersion('hexbin')`.

lubridate `r packageVersion('lubridate')`.

odbc `r packageVersion('odbc')`.

purrr `r packageVersion('purrr')`.

RSQLite `r packageVersion('RSQLite')`.

tidyr `r packageVersion('tidyr')`.

tidyverse `r packageVersion('tidyverse')`.
